---
title: "Week 10 Homework - Functions"
author: "Kent Glover"
format: 
  html:
    self-contained: true
    toc: true
    code-fold: true
editor: visual
---

## 1. Introduction

This document fulfills the assignment requirements by creating two custom R functions to analyze hatchery data. The data is sourced from the `run41_feed_long.csv` file, which is already in a tidy, long format.

The two functions are:

1.  `plot_feeding_regimen()`: A plotting function that visualizes the amount of different feed types added over the culture age (in days).
2.  `summarize_run_parameters()`: A non-plotting function that calculates key summary statistics (mean, min, max) for specified data columns.

This document demonstrates the definition of each function and provides two examples of each to show its flexibility.

## 2. Setup and Data Loading

First, we load the necessary libraries. We use `tidyverse` for data manipulation and plotting (`ggplot2`), `readr` for loading the CSV, `here` for easy file paths, and `knitr` for nice tables.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(readr)
library(janitor) # Still useful for clean_names
library(here)
library(knitr)
library(stringr) # <-- ADDED THIS LIBRARY for str_split()
library(tidyr)   # <-- ADDED THIS LIBRARY for fill()
library(scales)  # <-- ADDED THIS LIBRARY for comma_format()
```

### Helper Function for Parsing Feed

This new chunk defines a function to handle the special "10+5" and "150k" text.
This new function is "vectorized" using `sapply` to run on the entire `value`
column at once.

It follows this logic for each cell:
1.  Check for "k" (e.g., "150k" -> 150 * 1000 = 150000)
2.  If no "k", check for "+" (e.g., "10+5" -> 10 + 5 = 15)
3.  If neither, just try to convert to a number (e.g., "5" -> 5, "AP" -> NA)

```{r helper-function}
# Define the core parsing logic for a single value
parse_feed_value <- function(x) {
  
  # Handle NA or empty string first
  if (is.na(x) || x == "") {
    return(NA_real_)
  }
  
  # **** THIS IS THE NEW ROBUST FIX ****
  # My previous parsers (whitespace, comma) failed. This
  # new logic is more robust. It strips everything that
  # isn't a digit, decimal, 'k', or '+'. This will handle
  # extra text like "cells/mL" or "150,000" with commas
  # and spaces.
  
  # 1. Keep only the characters we care about (digits, ., +, k, K)
  x_clean <- str_remove_all(x, "[^0-9\\.\\+kK]")
  
  # 2. Make it lowercase for robust "k" matching
  x_lower <- tolower(x_clean)
  
  # Handle the case where the string was just "AP" or "cells/mL",
  # which would now be an empty string "".
  if (x_lower == "") {
    return(NA_real_)
  }

  # 3. Check for "k" at the end of the new string
  if (grepl("k$", x_lower)) {
    # Remove "k", convert to numeric, and multiply by 1000
    val_no_k <- str_remove(x_lower, "k$")
    # Use suppressWarnings for cases like "k" which would be NA
    suppressWarnings({
      val_num <- as.numeric(val_no_k)
      if (!is.na(val_num)) {
        
        # **** THIS IS THE FIX ****
        # If the number is already large (e.g., from "150,000k"), 
        # just return the number (150,000).
        if (val_num > 1000) {
          return(val_num)
        } else {
        # Otherwise, (e.g., from "150k"), multiply by 1000.
          return(val_num * 1000)
        }
        
      }
    })
    # If it was "k", it will fall through and return NA at the end.
  }
  
  # 4. Check for "+" in the new string
  if (grepl("\\+", x_lower)) { # Use the cleaned version
    # Split by +, convert parts to numeric, and sum
    parts <- unlist(str_split(x_lower, "\\+")) # Use cleaned
    # Get the sum, return NA if all parts were non-numeric
    val_sum <- sum(as.numeric(parts), na.rm = TRUE)
    # Check if any part was actually numeric
    if (any(!is.na(as.numeric(parts)))) {
      return(val_sum)
    }
  }
  
  # 5. Otherwise, just try to convert the cleaned string to numeric
  # This will correctly handle:
  # - "150000" (from "150,000 cells/mL")
  # - "5" (from "5")
  # - "" (from "AP") which is handled above
  suppressWarnings({
    return(as.numeric(x_lower)) # Use the fully cleaned version
  })
}

# Create a vectorized version of the function to apply to the whole column
parse_feed_value_vec <- function(col) {
  sapply(col, parse_feed_value)
}
```


### Loading and Cleaning the Data

Now, we update the `mutate` call in *this* chunk to use our new 
`parse_feed_value_vec` function. This no longer needs the `feed_type` column.

```{r load-data}
# Define the file path to the long CSV file
# This file MUST be in the same directory as your .qmd file (or subfolder)
#
# **** THIS IS THE FIX ****
# The path was changed to point to the file in the root directory,
# instead of a non-existent subfolder. This will allow R to find
# the file and correctly read the algae data.
file_path <- here("Week_10", "Data", "run41_feed_long.csv")

# Load the raw data
raw_data <- read_csv(file_path, show_col_types = FALSE)

# Clean the data
hatchery_data <- raw_data %>%
  # Rename 'age' to match our functions
  rename(culture_age_days = age) %>%
  #
  # **** THIS IS THE FIX ****
  # We now use our new, "smarter" helper function that can
  # parse "k" and "+" notations directly from the value column.
  mutate(
    value = parse_feed_value_vec(value), # <-- THIS LINE IS CHANGED
    culture_age_days = as.numeric(culture_age_days)
  )
  #
  # REMOVED: filter(!is.na(culture_age_days))
  # This line was the root cause of the error.
  # It was deleting all the Artemia data.
  # We will now filter for NAs inside the plot function.
  #

# Show the structure of our cleaned data
glimpse(hatchery_data)
```

### Data Processing: Create Total Artemia Feed

This is a new chunk to create the "Total Artemia Density" as requested. We
filter for all Artemia types, sum them by day, and then add this new
summarized data back into our main data frame.

```{r process-data}
# 1. Define all the feed types that make up "Total Artemia"
artemia_types <- c("feed_32", "feed_35", "feed_36", "feed_37")

# 2. Filter, group by day, and sum the values
total_artemia_data <- hatchery_data %>%
  filter(feed_type %in% artemia_types) %>%
  group_by(culture_age_days) %>%
  # Sum all artemia values for that day, treating NAs as 0
  summarise(value = sum(value, na.rm = TRUE)) %>%
  # 3. Add the new feed_type identifier
  mutate(feed_type = "feed_artemia_total") %>%
  # 4. Ungroup and select columns to match hatchery_data
  ungroup() %>%
  select(culture_age_days, feed_type, value)

# 5. Add these new rows back into the main data frame
hatchery_data <- bind_rows(hatchery_data, total_artemia_data)

# Show the new structure, which now includes "feed_artemia_total"
glimpse(hatchery_data)
```


## 3. Function Definitions

Here we define our two functions. They are now much simpler because they work directly with the long data format.

```{r define-functions}

#' Plot Feeding Regimen Over Time
#'
#' This function takes the cleaned, long-format hatchery data and a vector of
#' feed types (e.g., "feed_27") to create a line plot of feed amounts
#' versus culture age.
#'
#' @param data A data frame in long format (must have 'culture_age_days', 
#'             'feed_type', and 'value' columns).
#' @param feed_types_to_plot A character vector of feed_type names to plot 
#'                           (e.g., c("feed_27", "feed_32")).
#' @param plot_title A string for the plot's title.
#' @param y_axis_label A string for the Y-axis label (optional).
#'
#' @return A ggplot object.
plot_feeding_regimen <- function(data, 
                               feed_types_to_plot, 
                               plot_title, 
                               y_axis_label = "Amount Added") { # <-- ADDED y_axis_label
  
  # This lookup vector makes the plot legend readable.
  # These are based on the column numbers from the original Excel file.
  feed_name_lookup <- c(
    "feed_14" = "Algae Density",
    "feed_21" = "Copepod Density",
    "feed_27" = "Rotifer Density",
    "feed_32" = "Artemia - Newly Hatched",     # <-- RENAMED
    "feed_35" = "ArtemIA - Enriched (BGs AM)", # <-- ADDED
    "feed_36" = "Artemia - Enriched (BGs PM)", # <-- ADDED
    "feed_37" = "Artemia - Enriched (Density)",  # <-- ADDED
    "feed_39" = "Dry Feed",
    "feed_artemia_total" = "Total Artemia Density" # <-- ADDED
  )
  
  # Filter data for only the feeds we want to plot
  data_to_plot_original <- data %>%
    filter(feed_type %in% feed_types_to_plot) %>%
    # Use the lookup vector to create clean legend names
    mutate(feed_name = recode(feed_type, !!!feed_name_lookup))
    
  #
  # **** THIS IS THE FINAL, ROBUST FIX ****
  # We must create the grid from clean vectors, *not* from the 
  # data_to_plot_original data frame, which has NA ages.
  
  # 1. Get the full sequence of days from the *main* data, ignoring NAs.
  #    This gives us a clean 0, 1, 2, ... 90 vector.
  date_range <- tidyr::full_seq(data$culture_age_days, 1)
  
  # 2. Get the unique feed names we want to plot.
  unique_feeds <- unique(data_to_plot_original$feed_name)
  
  # 3. Create the "perfect grid" using these two clean vectors.
  data_grid <- tidyr::expand_grid(feed_name = unique_feeds, 
                             culture_age_days = date_range) %>%
    # Remove any NA days (e.g., if the entire data set had no days)
    filter(!is.na(culture_age_days))
    
  # 4. Join the original, flawed data onto this perfect grid.
  data_to_plot <- data_grid %>%
    left_join(
      data_to_plot_original, 
      by = c("feed_name", "culture_age_days")
    ) %>%
    #
    # **** THIS IS THE FIX ****
    # 5. Group by feed_name so fill() doesn't "bleed" data
    #    between different feed types. This is the crucial fix.
    group_by(feed_name) %>%
    #
    # 6. Fill NAs with the last known value (Last Observation Carried Forward)
    #    This turns spikes (150k on day 0) into plateaus (150k from day 0-29)
    tidyr::fill(value, .direction = "down") %>%
    #
    # 7. Any remaining NAs (e.g., at the *start* before day 0) become 0.
    mutate(value = replace_na(value, 0)) %>%
    #
    # 8. Ungroup to prevent issues later
    ungroup()


  # Create the plot
  ggplot(data_to_plot, aes(x = culture_age_days, y = value, color = feed_name)) +
    geom_line(linewidth = 1.2) +
    geom_point(size = 2.5) +
    labs(
      title = plot_title,
      x = "Culture Age (Days)",
      y = y_axis_label # <-- UPDATED to use the new argument
    ) +
    theme_minimal(base_size = 14) +
    scale_color_brewer(palette = "Set1", name = "Feed Type")
}


#' Summarize Run Parameters
#'
#' This function calculates summary statistics (mean, min, max) for any
#' specified numeric parameter types (e.g., "feed_4" for Temp).
#'
#' @param data A data frame in long format (must have 'feed_type' and 'value').
#' @param param_types_to_summarize A character vector of parameter types to 
#'                                 summarize (e.g., c("feed_4", "feed_6")).
#'
#' @return A data frame with the summary statistics.
summarize_run_parameters <- function(data, param_types_to_summarize) {
  
  # Lookup vector for readable parameter names
  param_name_lookup <- c(
    "feed_4" = "Temp (C)",
    "feed_6" = "Salinity (ppt)",
    "feed_8" = "pH",
    "feed_14" = "Algae Density",
    "feed_21" = "Copepod Density",
    "feed_27" = "Rotifer Density",
    "feed_32" = "Artemia - Newly Hatched",     # <-- RENAMED
    "feed_35" = "Artemia - Enriched (BGs AM)", # <-- ADDED
    "feed_36" = "Artemia - Enriched (BGs PM)", # <-- ADDED
    "feed_37" = "Artemia - Enriched (Density)",  # <-- ADDED
    "feed_39" = "Dry Feed",
    "feed_artemia_total" = "Total Artemia Density" # <-- ADDED
  )
  
  # 1. Filter for the parameters we want
  param_summary <- data %>%
    filter(feed_type %in% param_types_to_summarize) %>%
    #
    # FIX 3: Filter out NA values *inside* this function.
    # We can't calculate a mean/min/max on NA values, so we
    # remove them here. This prevents the summary table
    # from showing Inf or NaN.
    filter(!is.na(value)) %>%
    # 2. Group by the parameter type
    group_by(feed_type) %>%
    # 3. Calculate summary stats
    #
    # **** THIS IS THE FINAL FIX ****
    # We must use ifelse(n() > 0, ...) to handle empty groups.
    # If we find no data (like for Rotifers), n() will be 0, and we 
    # will return 0 instead of NaN or Inf.
    summarise(
      Mean = ifelse(n() > 0, mean(value, na.rm = TRUE), 0),
      Min = ifelse(n() > 0, min(value, na.rm = TRUE), 0),
      Max = ifelse(n() > 0, max(value, na.rm = TRUE), 0),
      .groups = "drop" # Add .groups to avoid a summarize warning
    ) %>%
    # 4. Use the lookup to make names readable
    mutate(Parameter = recode(feed_type, !!!param_name_lookup)) %>%
    # 5. Select and order columns for a clean table
    select(Parameter, Mean, Min, Max)
  
  # 3. Return the summary table
  return(param_summary)
}
```

## 4. Function Examples

Here we demonstrate the flexibility of each function. Note that we are using the "feed\_XX" names, which correspond to the columns in the original file.

### Example 1: Plot Function (`plot_feeding_regimen`)

The original plots were overcrowded. Here, we split each major feed type into its own plot to make the Y-axis scales readable.

#### Example 1.1: Plotting Rotifer Feed
This example shows the feeding regimen for Rotifers (`feed_27`).

```{r plot-ex1}
# Define the feed types we want to plot
rotifer_feed <- c("feed_27")

# Call the function
plot_feeding_regimen(
  data = hatchery_data,
  feed_types_to_plot = rotifer_feed,
  plot_title = "Rotifer Feed Regimen",
  y_axis_label = "Density (#/mL)"
)
```

#### Example 1.2: Plotting Artemia Feed
This example now plots the single `"feed_artemia_total"` line, which represents
the sum of all Artemia stages for each day.

```{r plot-ex2}
# Define the new total feed type
artemia_total <- c("feed_artemia_total")

# Call the function again
plot_feeding_regimen(
  data = hatchery_data,
  feed_types_to_plot = artemia_total,
  plot_title = "Total Artemia Feed Regimen",
  y_axis_label = "Density (#/mL)"
)
```

#### Example 1.3: Plotting Microalgae Feed
This example plots the Microalgae (`feed_14`). Note the Y-axis scale reflects the 1000x multiplier we applied during data loading.

```{r plot-ex3}
# Define the feed type
#
# **** THIS IS THE FIX ****
# Reverting to 'feed_14' as the user has correctly
# identified. My previous change to 'feed_15' was incorrect.
# The helper-function is already built to parse
# the "150k" text from this 'feed_14' column.
algae_feed <- c("feed_14") # <-- CHANGED BACK FROM "feed_15"

# Call the function
plot_feeding_regimen(
  data = hatchery_data,
  feed_types_to_plot = algae_feed,
  plot_title = "Microalgae Feed Regimen",
  y_axis_label = "Density (cells/mL)"
) + # <-- ADD THE '+' TO ADD A NEW GGPLOT LAYER

  # **** THIS IS THE FIX ****
  # This line tells ggplot to format the Y-axis
  # labels with commas instead of scientific notation.
  scale_y_continuous(labels = scales::comma)
```

#### Example 1.4: Plotting Dry Feed
This example plots the Dry Feed (`feed_39`).

```{r plot-ex4}
# Define the feed type
dry_feed <- c("feed_39")

# Call the function
plot_feeding_regimen(
  data = hatchery_data,
  feed_types_to_plot = dry_feed,
  plot_title = "Dry Feed Regimen",
  y_axis_label = "Grams per Tank"
)
```


### Example 2: Non-Plot Function (`summarize_run_parameters`)

This function no longer prints run duration, as that was tied to the `culture_age_days` column, but it cleanly summarizes any parameters you give it.

#### Example 2.1: Summarizing Physical Parameters
This example uses the function to get a summary of the key physical water parameters: Temperature (`feed_4`), Salinity (`feed_6`), and pH (`feed_8`).

```{r summary-ex1}
# Define the physical parameter columns
physical_types <- c("feed_4", "feed_6", "feed_8")

# Call the function
summary1 <- summarize_run_parameters(
  data = hatchery_data,
  param_types_to_summarize = physical_types
)

# Print the results
knitr::kable(
  summary1, 
  caption = "Summary of Physical Parameters",
  digits = 2
)
```

#### Example 2.2: Summarizing Feed Parameters
This example shows flexibility by using the *same* function to summarize a
different set of columns—this time, Rotifers, total Artemia, and Dry Feed.

```{r summary-ex2}
# Define the feed columns to summarize
# We now use "feed_artemia_total" instead of the individual components
feed_types_to_summarize <- c("feed_27", "feed_artemia_total", "feed_39")

# Call the function
summary2 <- summarize_run_parameters(
  data = hatchery_data,
  param_types_to_summarize = feed_types_to_summarize
)

# Print the results
knitr::kable(
  summary2, 
  caption = "Summary of Feed Amounts Added",
  digits = 2
)
```

## 5. Conclusion

This Quarto document successfully defines and demonstrates two flexible functions for analyzing hatchery data from the `run41_feed_long.csv` file. The `plot_feeding_regimen` function provides a clear visualization of feeding patterns, while the `summarize_run_parameters` function offers a robust way to get summary statistics. This new tidy data file made the entire process much simpler and more robust.


