---
title: "Tidy Tuesday Week 14: Statistical Performance Indicators"
author: "Kent Glover"
format: 
  html:
    embed-resources: true
    theme: cosmo
---

## 1. Introduction and Setup

This week we are analyzing the World Bank's **Statistical Performance Indicators (SPI)**. The goal is to see how country statistical systems have improved (or declined) over time.

For my "New Thing" this week, I am creating a **Dumbbell Chart** to visualize the change in scores between two specific years.

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(here)

#set a good theme
theme_set(theme_minimal(base_size = 12))
```

## 2\. Load Data

I am pulling the data directly from the TidyTuesday repository.

```{r}
#| message: false
#| warning: false

#read directly from Github
spi_indicators <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-11-25/spi_indicators.csv')


#quick look
glimpse(spi_indicators)
```

## 3\. Data Wrangling

To create a clean visualization, I need to subset the data.

1.  I will focus on the **"Latin America & Caribbean"** region to keep the number of countries manageable.
2.  I will compare the `overall_score` between the years **2019** and **2023** to see recent changes.
3.  I need to pivot the data `wider` so I have separate columns for 2019 and 2023 to draw the "dumbbell" lines between them.

<!-- end list -->

```{r}
#Filter for region and years of interest
dumbbell_data <- spi_indicators %>%
  filter(region == "Latin America & Caribbean") %>%
  filter(year %in% c(2019, 2023)) %>%
  select(country, year, overall_score) %>%
  #pivot wider to get start and end points for the segments
  pivot_wider(names_from = year, values_from = overall_score, names_prefix = "year_") %>%
  #remove countries that don't have data for both years
  drop_na() %>%
  #calculate the change for sorting and coloring
  mutate(change = year_2023 - year_2019,
         direction = ifelse(change > 0, "Improved", "Declined")) %>%
  # reorder country factor by 2023 score so the plot is sorted
  mutate(country = fct_reorder(country, year_2023))

head(dumbbell_data)
```

## 4\. Visualization: The "New Thing"

::: callout-note

## My "New Thing": The Dumbbell Chart

Instead of using a clustered bar chart, I learned how to build a **Dumbbell Chart** using `geom_segment()` and `geom_point()`.

This is considered a best practice for "Before vs. After" comparisons because it reduces visual clutter and clearly shows the *distance* and *direction* of change.
:::

```{r plot-dumbbell, fig.width=10, fig.height=8}
#create plot
p_dumbbell <- ggplot(dumbbell_data) +
  # 1.The "bar" of the dumbbell (the segment connecting the years)
  geom_segment(aes(x = year_2019, xend = year_2023, 
                   y = country, yend = country), 
               color = "grey60", size = 1) +
  
  # 2.Point for 2019 (Start)
  geom_point(aes(x = year_2019, y = country), 
             color = "grey60", size = 3) +
  
  # 3.Point for 2023 (End) colored by improvement status
  geom_point(aes(x = year_2023, y = country, color = direction), 
             size = 3) +
  
  #custom colors for the direction
  scale_color_manual(values = c("Improved" = "#2E8B57", "Declined" = "#CD5C5C")) +
  
  #better labels
  labs(title = "Statistical Performance in Latin America (2019 vs 2023)",
       subtitle = "Comparing the World Bank's Statistical Performance Index (SPI) Overall Score.\nGreen dots indicate the 2023 score.",
       x = "SPI Overall Score (0-100)",
       y = NULL,
       color = "Status",
       caption = "Data: World Bank via TidyTuesday") +
  
  #theme adjustments for cleanliness
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(), #remove horizontal grid lines
    plot.title = element_text(face = "bold", size = 16)
  )

#display
p_dumbbell

#Save Output
ggsave(here("Tidy_Tuesday", "Week_14", "Outputs", "spi_dumbbell_plot.png"), 
       p_dumbbell, width = 10, height = 8)
```

## 5\. Summary

By using the **Dumbbell Chart**, we can easily identify that while most countries in the region improved their statistical capacity (like **Mexico** and **Chile**), a few saw declines. This visualization is much cleaner than trying to read two bars side-by-side for each country.