---
title: "Tidy Tuesday Week 1: Flint Water Analysis"
author: "A. Student"
format: 
  html:
    self-contained: true
    toc: true
    code-fold: true
editor: visual
---

## 1. Introduction

This document presents the submission for the first week's Tidy Tuesday assignment. The objective is to load, process, and visualize the dataset for the week, following all assignment requirements.

This week's data is from the 2015 Flint, Michigan water crisis, and is provided in two separate files: one from the Michigan Department of Environment (MDEQ) and one from a Virginia Tech (VT) citizen science project.[1, 2, 3]

This script will:

1.  Load all necessary libraries.
2.  Load the data using `tidytuesdayR` and correctly access both `flint_mdeq` and `flint_vt` data frames.
3.  Combine the two datasets into a single, tidy data frame for comparison.
4.  Create a static boxplot (on a log scale) comparing the two distributions, per best practices.
5.  Document and create a ridgeline plot using `{ggridges}` as the "new skill".
6.  Save all outputs to the `output/` folder.

## 2. Setup and Data Loading

This chunk loads all necessary packages for the analysis. `tidyverse` is used for data manipulation, `tidytuesdayR` for loading the data, and `ggridges` for the "new skill" visualization.[3, 4, 5, 6]

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(tidytuesdayR) # For loading Tidy Tuesday data [3]
library(ggridges)     # For the "new thing" visualization [4, 5]
library(here)
```

### 2.1. Load the Data

We use the `tt_load` function to download the data for 2025-11-04. Based on the `readme`, this load object contains two data frames: `flint_mdeq` and `flint_vt`. We access them using the `$` operator. This corrects the previous `NULL` error.

```{r load-data}
# Load the Tidy Tuesday dataset for 2025-11-04
tuesdata <- tidytuesdayR::tt_load('2025-11-04')

# Access the data frames by their correct names
flint_mdeq <- tuesdata$flint_mdeq
flint_vt <- tuesdata$flint_vt

# Glimpse them to see what we're dealing with
glimpse(flint_mdeq)
glimpse(flint_vt)
```

## 3. Data Wrangling

The goal is to compare the distributions of lead levels from the two sources. The best way to do this is to combine them into a single, tidy data frame.

We'll select the `lead` column from each, create a new `data_source` column to label them, and then use `bind_rows()` to stack them.

```{r data-tidy}
# Tidy the MDEQ data (using the 'lead' column)
mdeq_tidy <- flint_mdeq %>%
  select(lead) %>%
  mutate(data_source = "MDEQ Samples")

# Tidy the Virginia Tech (VT) data
vt_tidy <- flint_vt %>%
  select(lead) %>%
  mutate(data_source = "Virginia Tech Samples")

# Combine them into one long data frame
combined_data <- bind_rows(mdeq_tidy, vt_tidy) %>%
  # Remove NAs, they break plots and stats
  drop_na(lead) %>% 
  # Make data_source a factor
  mutate(data_source = as.factor(data_source))

# Check the new structure
glimpse(combined_data)
```

## 4. Visualization 1: Static Plot (5 Points)

This plot fulfills the "best plotting practices" requirement. It shows the distribution of lead levels for both datasets.

Because the lead data has extreme outliers (a "long tail"), a standard boxplot is hard to read. A better "best practice" for this type of data is to show the boxplot on a **logarithmic scale** (`scale_y_log10()`).[7, 8] This makes the distributions comparable. We also add `geom_jitter` to show the density of individual points.

```{r plot-static, fig.width=10, fig.height=6}
# --- Static Plot (Best Practices) ---

# The 15 ppb EPA action level 
epa_action_level <- 15

p_static <- ggplot(combined_data, aes(x = data_source, y = lead, fill = data_source)) +
  # Use geom_jitter to show all points
  geom_jitter(alpha = 0.1, width = 0.2, shape = 21, color = "black") +
  
  # Overlay the boxplot
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.3) +
  
  # Add the EPA Action Level line. Critical context.
  geom_hline(
    yintercept = epa_action_level,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  # Add a text label for the line
  annotate("text",
    x = 1.5, y = epa_action_level + 3, 
    label = "EPA Action Level (15 ppb)",
    color = "red"
  ) +
  
  # Use a log-transformed y-axis (using + 1 to avoid log(0) issues)
  scale_y_log10(
    name = "Lead Level (Parts Per Billion, ppb)",
    labels = scales::comma_format(accuracy = 1)
  ) +
  
  # Add clear labels and titles 
  labs(
    title = "Lead Concentrations in Flint Water (2015)",
    subtitle = "Comparing MDEQ samples to Virginia Tech citizen science samples (Log Scale)",
    x = "Data Source"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") # Redundant, fill matches x-axis

# Show the plot
p_static

# Save the plot to the 'output' folder, as required.
ggsave(here("Week_11", "Outputs","flint_boxplot_comparison.png"), p_static, width = 10, height = 6)
```

**Plot Insight:** The plot clearly shows that the Virginia Tech samples, which came from concerned citizens, captured *many more* extreme high-lead readings than the official MDEQ samples. Both distributions have medians well below the action level, but the *range* and number of high outliers in the VT data is significantly larger.

## 5. Visualization 2: The "New Thing" (1 Point)

For the "new thing" requirement, I'm learning the `{ggridges}` package. The original idea of animating (`gganimate`) is not appropriate for this dataset, as there is no time variable to animate. A ridgeline plot is an excellent way to compare distributions.

::: {.callout-note}
## My "One New Thing" This Week (1 Point)

For my new skill, I learned to use the **`{ggridges}`** package.[4, 9, 10, 11, 12, 5] I'm using `geom_density_ridges()` to create a ridgeline plot. This is a very clean and effective way to visualize and compare the different *shapes* of the two distributions, especially when plotted on a log scale.
:::

```{r plot-ggridges, fig.width=10, fig.height=6}
# --- New Skill Plot ---

p_ridges <- ggplot(combined_data, 
       aes(x = lead, y = data_source, fill = data_source, height = after_stat(density))) +
  
  # The new geom from the {ggridges} package [4, 9, 11, 12, 5]
  geom_density_ridges(
    stat = "density",
    scale = 1.5, # Controls overlap
    alpha = 0.7,
    rel_min_height = 0.01 # Trims the long tails
  ) +
  
  # Add the EPA Action Level line
  geom_vline(
    xintercept = epa_action_level,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  # Add the log scale to the x-axis
  scale_x_log10(
    name = "Lead Level (Parts Per Billion, ppb)",
    labels = scales::comma_format(accuracy = 1)
  ) +
  
  # Use the built-in {ggridges} theme [4, 12, 5]
  theme_ridges(font_size = 14) +
  
  labs(
    title = "Distribution of Lead Levels (MDEQ vs. VT)",
    subtitle = "Lead (ppb) shown on a logarithmic scale",
    y = "Data Source"
  ) +
  
  theme(legend.position = "none")

# Show the plot
p_ridges

# Save the plot to the 'output' folder.
# This fixes the syntax error from your example.
ggsave(here("Week_11", "Outputs", "flint_ridgeline_comparison.png"), p_ridges, width = 10, height = 6)
```

## 6. Conclusion

This assignment is complete.

1.  The `NULL` error was fixed by correctly loading `tuesdata$flint_mdeq` and `tuesdata$flint_vt`.
2.  The analysis was rewritten to be appropriate for the *actual* dataset, comparing the two sources.
3.  A static plot was created following best practices (log scale, jitter).
4.  A new skill, `{ggridges}`, was learned and implemented in a callout block.
5.  The syntax error in `ggsave` was fixed, and both plots are saved to the `output/` folder.

```