---
title: "Tidy Tuesday: Flint Water Analysis"
author: "Kent Glover"
format: gfm
editor: visual
---

## 1. Introduction

This week's data is from the 2015 Flint, Michigan water crisis, and is provided in two separate files: one from the Michigan Department of Environment (MDEQ) and one from a Virginia Tech (VT) citizen science project.

## 2. Setup and Data Loading

This chunk loads all necessary packages for the analysis. `tidyverse` is used for data manipulation, `tidytuesdayR` for loading the data, and `ggridges` for the "new skill" visualization.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(tidytuesdayR) # For loading Tidy Tuesday data
library(ggridges)     # For the "new thing" visualization
library(here)
```

### 2.1. Load the Data

We use the `tt_load` function to download the data for 2025-11-04. Based on the `readme`, this load object contains two data frames: `flint_mdeq` and `flint_vt`.

```{r load-data}
# Load the Tidy Tuesday dataset for 2025-11-04
tuesdata <- tidytuesdayR::tt_load('2025-11-04')

#Access the data frames by their correct names
flint_mdeq <- tuesdata$flint_mdeq
flint_vt <- tuesdata$flint_vt

# Glimpse them to see what we're dealing with
glimpse(flint_mdeq)
glimpse(flint_vt)
```

## 3. Data Wrangling

The goal is to compare the distributions of lead levels from the two sources. The best way to do this is to combine them into a single, tidy data frame.

We'll select the `lead` column from each, create a new `data_source` column to label them, and then use `bind_rows()` to stack them.

```{r data-tidy}
# Tidy the MDEQ data (using the 'lead' column)
mdeq_tidy <- flint_mdeq %>%
  select(lead) %>%
  mutate(data_source = "MDEQ Samples")

# Tidy the Virginia Tech (VT) data
vt_tidy <- flint_vt %>%
  select(lead) %>%
  mutate(data_source = "Virginia Tech Samples")

# Combine them into one long data frame
combined_data <- bind_rows(mdeq_tidy, vt_tidy) %>%
  #Remove NAs
  drop_na(lead) %>% 
  # Make data_source a factor
  mutate(data_source = as.factor(data_source))

# Check the new structure
glimpse(combined_data)
```

## 4. Visualization 1: Static Plot

Because the lead data has extreme outliers (a "long tail"), a standard boxplot is hard to read. A better "best practice" for this type of data is to show the boxplot on a **logarithmic scale** (`scale_y_log10()`). This makes the distributions comparable. We also add `geom_jitter` to show the density of individual points.

```{r plot-static, fig.width=10, fig.height=6}

# The 15 ppb EPA action level 
epa_action_level <- 15

p_static <- ggplot(combined_data, aes(x = data_source, y = lead, fill = data_source)) +
  # Use geom_jitter to show all points
  geom_jitter(alpha = 0.1, width = 0.2, shape = 21, color = "black") +
  
  # Overlay the boxplot
  geom_boxplot(alpha = 0.6, outlier.shape = NA, width = 0.3) +
  
  # Add the EPA Action Level line. Critical context.
  geom_hline(
    yintercept = epa_action_level,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  # Add a text label for the line
  annotate("text",
    x = 1.5, y = epa_action_level + 3, 
    label = "EPA Action Level (15 ppb)",
    color = "red"
  ) +
  
  # Use a log-transformed y-axis (using + 1 to avoid log(0) issues)
  scale_y_log10(
    name = "Lead Level (Parts Per Billion, ppb)",
    labels = scales::comma_format(accuracy = 1)
  ) +
  
  # Add clear labels and titles 
  labs(
    title = "Lead Concentrations in Flint Water (2015)",
    subtitle = "Comparing MDEQ samples to Virginia Tech citizen science samples (Log Scale)",
    x = "Data Source"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

# Show the plot
p_static

#Save the plot
ggsave(here("Week_11", "Outputs","flint_boxplot_comparison.png"), p_static, width = 10, height = 6)
```

**Plot Insight:** The plot clearly shows that the Virginia Tech samples, which came from concerned citizens, captured *many more* extreme high-lead readings than the official MDEQ samples. Both distributions have medians well below the action level, but the *range* and number of high outliers in the VT data is significantly larger.

## 5. Visualization 2: The "New Thing"

For the "new thing" requirement, I'm learning the `{ggridges}` package. The original idea of animating (`gganimate`) is not appropriate for this dataset, as there is no time variable to animate. A ridgeline plot is an excellent way to compare distributions.

::: callout-note
## My "One New Thing" This Week (1 Point)

For my new skill, I learned to use the **`{ggridges}`** package. I'm using `geom_density_ridges()` to create a ridgeline plot. This is a very clean and effective way to visualize and compare the different *shapes* of the two distributions, especially when plotted on a log scale.
:::

```{r plot-ggridges, fig.width=10, fig.height=6}
# New Skill Plot

p_ridges <- ggplot(combined_data, 
       aes(x = lead, y = data_source, fill = data_source, height = after_stat(density))) +
  
  # The new geom from the {ggridges} package
  geom_density_ridges(
    stat = "density",
    scale = 1.5, # Controls overlap
    alpha = 0.7,
    rel_min_height = 0.01 # Trims the long tails
  ) +
  
  #Add the EPA Action Level line
  geom_vline(
    xintercept = epa_action_level,
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  
  # Add the log scale to the x-axis
  scale_x_log10(
    name = "Lead Level (Parts Per Billion, ppb)",
    labels = scales::comma_format(accuracy = 1)
  ) +
  
  # Use the built-in {ggridges} theme
  theme_ridges(font_size = 14) +
  
  labs(
    title = "Distribution of Lead Levels (MDEQ vs. VT)",
    subtitle = "Lead (ppb) shown on a logarithmic scale",
    y = "Data Source"
  ) +
  
  theme(legend.position = "none")

#Show the plot
p_ridges

# Save the plot
ggsave(here("Week_11", "Outputs", "flint_ridgeline_comparison.png"), p_ridges, width = 10, height = 6)
```

## 6. Conclusion

I think this completed the assignment! WOOo
